---
output:
  html_document: default
  word_document: default
  pdf_document: default
---
# Introduction {#intro}

Research methods build the base for innovation in many areas. As one domain recently learning engineering looks at the challenge of continuous renewal and innovation in educational technologies and the evluation of these innovations.

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

This is also a reference to [@kobsa2001personalised] this is important.

```{r}
# Loop to print numbers from 1 to 10
for (i in 1:10) {
  print(i)
}
```

```{r}
install.packages("DiagrammeR")
library(DiagrammeR)

# Create a blank diagram
flowchart <- create_graph()

# Add nodes to the diagram
flowchart <- add_node(flowchart, "Start", shape = "rectangle")
flowchart <- add_node(flowchart, "Process 1", shape = "rectangle")
flowchart <- add_node(flowchart, "Process 2", shape = "rectangle")
flowchart <- add_node(flowchart, "End", shape = "rectangle")

# Add edges to connect the nodes
flowchart <- add_edge(flowchart, from = "Start", to = "Process 1")
flowchart <- add_edge(flowchart, from = "Process 1", to = "Process 2")
flowchart <- add_edge(flowchart, from = "Process 2", to = "End")

# Render and display the flowchart diagram
render_graph(flowchart)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
